---
title: "A link function for logistic regressions"
author: "Jürgen Wilbert"
abstract: A brief explanation of the link function in logistic regression.
date: 2023/06/22
format:
  html:
    page-layout: full
    df-print: kable
    toc: true
    toc-location: right
    self-contained: false
    code-fold: true
categories:
  - regression
  - statistics
  - link
  - logistic
editor: 
  markdown: 
    wrap: 72
---

## Problem and solution

If you estimate a dichotomous outcome variable $y$ ($y$ can take the
values 0 *or* 1) with a standard *ols* (ordinary least squares)
regression the distribution of $y$ does not meet *ols* regression
assumptions. Specifically, predicted values can be outside $\{0, 1\}$
and are not continuous and normal distributed. A common solution for
this problem is to apply a generalization of the linear regression model
which is (surprise, surprise) called the generalized linear model (GLM).
One aspect of the generalized linear model is that the linear
combination of the predictors (linear because we assume that the added
up predictors are interval scaled) are linked to the distribution of the
outcome variable via a function which is called: the link function
(uhhh).

It is helpful to keep in mind that for dichotomous variables, we
estimate/predict the probabilities that $y=1$. That is, the predicted
values are probabilities of $[0, 1]$ (a continuous interval *between* 0
and 1) and not $\{0, 1\}$ (either 0 *or* 1). Therefore, we are looking
for a link function that turns the linear combined predictors into a
probability estimation. In case of a logistic regression, we need a
formula, that turns continuous values from $[-\infty, +\infty]$ to a
probability (where $0$ corresponds to a 50% probability) and vise-versa,
a probability into a continuous scale with values from
$[-\infty, +\infty]$. This transformation is called a logit
transformation and the distribution of the transformed predicted $y$
values is a logistic distribution. This kind of function (i.e. a
function that links the distribution of linear combinations of the
predictors to the distribution of the criteria in a regression) is
called a link function.

## The logit transformation

The formula to turn probabilities into continuous values (here logits)
is:

$$logit(y_i)=log(\frac{P(y_i=1)}{1-P(y_i=0)})=log(odds)$$

Here is the function plot for illustration with probabilities on x and
logits on y:

```{r}
#| fig-cap: Logit/logistic function

# identical: plot(function(x) qlogis(x))
plot(function(x) log(x/(1-x)))
```

The formula to turn continuous values (here logits) into probabilities
is:

$$P(y_i=1)=\frac{exp(x)}{1 + exp(x)}$$

Here is the function plot for illustration with logits on x and
probablities on y:

```{r}
#| fig-cap: Distribution of the logistic function (invers logistic) 
# identica: plot(function(x) plogis(x), -6,6)
plot(function(x) exp(x)/(1 + exp(x)), from = -6, to = 6)
```

## What are logits?

Logits are the logarithm of the odds of a probability $log(Odds)$. Odds
is a way to represent the probability of an event: If the odds are four
against one ($4\over1$), than out of 5 events, 4 have one category of
outcome and 1 has an opposite outcome (for example, if a horse runs five
races and we expect that it will win four races and loose one, the odds
are $4\over1$). If the odds are \< 1, the probability of an event is
below 50%. For example, if the odds are 0.25 = $\frac{1}{4}$ we expect
that the horse will win one out of five races.

::: callout-caution
Caution: Probabilities and odds can be confused easily but are not
identical. A probability of $1\over4$ would index that a specific
outcome happens in 1 *out of* 4 events, whereas odds of $1\over4$ would
index that an event has the probability for two possible outcomes of 1
*against* 4. For this example, odds of $1\over4$ equals a probability of
$1\over5$ equals $0.20$.
:::

The odds are calculated as:

$$Odds(y_i)=\frac{P(y_i=1)}{1-P(y_i=0)}$$

For example, if we have a probability of 25%, the odds are:

$$Odds=\frac{0.25}{0.75}={1\over3}$$\
The formula for calculating logits is:

$$logit(y_i)=log(\frac{P(y_i=1)}{1-P(y_i=0)})=log(odds)$$

For example, if we have a 25% probability that is:

$$logit(P=0.25)=log(\frac{0.25}{0.75})=log({1\over3})=-1.098612$$

## Example

To be continued 😉

```{r}
#| include: false
#| fig-cap: Logit/logistic density function
plot(function(x) dlogis(x), -6,6)
```
