---
title: eRm
author: Jürgen Wilbert
date: 2021/05/31
abstract: Extended Rasch modelling
output:
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: yes
    df_print: default
    theme: journal
keywords: 
  - irt
  - Rasch
  - eRm
---

```{r results='asis', echo = FALSE}
cat("*Keywords: ", paste0(rmarkdown::metadata$keywords, collapse = ", "), "*"
)
```

```{r setup, include = FALSE}
library(tidyverse)
library(eRm)

knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6, echo = TRUE
)
```

# Prepare data

```{r}
dat <- readRDS("test.rds")

dat <- dat %>% 
  mutate(
    sum_d = rowSums(across(starts_with("D_"))),
    sum_a = rowSums(across(starts_with("A_")))
  )

dat_d <- dat %>% select(starts_with("D_"))
dat_a <- dat %>% select(starts_with("A_"))
```

# Fit model

```{r}
fit_d <- RM(dat_d)
fit_a <- RM(dat_a)
```

# Report item difficulty

```{r}
summary(fit_d)
summary(fit_a)
```

# Person parameters

## Extract

```{r}
pp_a <- person.parameter(fit_a)
pp_d <- person.parameter(fit_d)

dat$pp_d <- coef(pp_d) # add person parameter to raw data frame
dat$pp_a <- coef(pp_a) # add person parameter to raw data frame
```

Caution! Raw values of 0 and 15 (upper and lower limit) are between -Inf/+Inf and the upper/lower cut off and are estimated based on the distribution shape.

```{r}
plot(pp_d, main = "Dekodierungsleistung")
plot(pp_a, main = "Automatisierung")
```

```{r}
plotPImap(fit_d)
```



# Test Modelfit

## Visual insoection of item discrimination

```{r}
plotjointICC(fit_d, xlim = c(-5, 5))
plotjointICC(fit_a, xlim = c(-5, 5))
```


## Likelihood ratio tests

LRtests estimate the estimation "stability" across several subsamples of the data set.

A median split allows for cheking if the item-difficulty estimations are constant for the lower and upper part of the performance scale. This is actually a test for homoscedasticity.

```{r}
lr_d <- LRtest(fit_d, splitcr = "median")
lr_a <- LRtest(fit_a, splitcr = "median")

lr_d
lr_a

```

```{r}
plotGOF(lr_d, conf= list())
plotGOF(lr_a, ctrline= list())
```



# Wald Test

```{r}
Waldtest(fit_d)
Waldtest(fit_a)

```

## Item In-Fit / Out Fit

ItemFit interpretation:

- larger 2.0: Distorts or degrades the measurementsystem
- 1.5-2.0: Unproductive for construction of measurement, but not degrading
- 0.5-1.5: Productive for measurement
- <0.5: Lessproductive for measurement, but not degrading. May produce misleadingly good reliabilities and separations

```{r}
itemfit(pp_d)
itemfit(pp_a)
```

## Person In-Fit / Out Fit

Personfit interpretation:

- larger 2.0: Distorts or degrades the measurementsystem
- 1.5-2.0: Unproductive for construction of measurement, but not degrading
- 0.5-1.5: Productive for measurement
- <0.5: Lessproductive for measurement, but not degrading. May produce misleadingly good reliabilities and separations

```{r}
personfit(pp_d)
personfit(pp_a)
```

