---
title: "Contrasts in single-cases"
author: Jürgen Wilbert
date: 2022/09/27
abstract: Contrasts in regression modells
categories: 
  - contrasts
  - regression
  - statistics
draft: true
---

```{r}
#| include: false
library(scan)
library(scplot)
knitr::opts_chunk$set(comment = '')
```

# Example dataset

::: panel-tabset

## plot

```{r}
#| echo: false
scplot(exampleA1B1A2B2$Pawel) |> 
  add_statline("trendA", color = "red") |> 
  add_statline("mean", color ="red")
```

## table

```{r}
#| echo: false
export(exampleA1B1A2B2$Pawel)
```

:::

```{r}
#| echo: false
# dataset:
df <- as.data.frame(exampleA1B1A2B2$Pawel)
df$mt <- df$mt - df$mt[1]
```


```{r}
#| collapse: true

# mean of all phases

(means <- tapply(df$values, df$phase, mean))
(grand_mean <- mean(means))
```

```{r}
#| include: false

#???
cum_contr_mean <- matrix(c(-1,1/3,1/3,1/3, -0.5,-0.5,0.5,0.5, -1/3,-1/3,-1/3,1), ncol = 3)
colnames(cum_contr_mean) <- levels(df$phase)[-1]

cum_contr_mean

cum_contr_mean <- matrix(
  c(1/4,1/4,1/4,1/4, 
    -1,1/3,1/3,1/3, 
    -0.5,-0.5,0.5,0.5, 
    -1/3,-1/3,-1/3,1), ncol = 4)
cum_contr_mean
cum_contr_mean <- solve(t(cum_contr_mean))[,-1]
colnames(cum_contr_mean) <- levels(df$phase)[-1]
cum_contr_mean


```

```{r}
# Plm function reference:
plm(exampleA1B1A2B2$Pawel, contrast = "first", trend = FALSE, slope = FALSE) %>% coef() %>% .[,1]

plm(exampleA1B1A2B2$Pawel, contrast = "preceding", trend = FALSE, slope = FALSE) %>% coef() %>% .[,1]
```

# Contrasts

## Treatment contrast

*Compare mean of second to last phase against first phase (intercept)*

The Intercept is the mean of a reference phase (defaults to the first).\
The predictors are the differences from a phase to the reference (intercept).

```{r}
treatment <- contr.treatment(4)
colnames(treatment) <- c("B1vsA1","A2vsA1","B2vsA1")
treatment

lm(values~phase, data=df, contrasts = list(phase = treatment)) |> coef()

c(Intercept = 13.7, # phaseA1
  phaseB1vsA1 = 25.9 - 13.7, 
  phaseA2vsA1 = 14.7 - 13.7,
  phaseB2vsA1 = 27 - 13.7)
```

## Sum contrast

*Comparison from the first to the second last phase with the overall mean (Intercept is overall mean)*

The Intercept is the overall mean.\
The predictors are the differences from a phase mean to the overall mean.

```{r}
sum <- contr.sum(4)
colnames(sum) <- c("A1vsMean","B1vsMean","A2vsMean")
sum

lm(values~phase, data=df, contrasts = list(phase = sum)) |> coef()

c(Intercept = grand_mean,
  phaseA1vsMean = 13.7 - grand_mean, 
  phaseB1vsMean = 25.9 - grand_mean, 
  phaseA2vsMean = 14.7 - grand_mean)
```

## Helmert contrast

*Compares from the second to the last phase with the mean of all preceding phases.*

The Intercept is the overall mean.\
The predictors are the mean differences from a phase mean to the mean of all preceding phases.

```{r}
helmert <- contr.helmert(4)
colnames(helmert) <- c("B1vsA1", "A2vsA1_B1", "B2vsA1_B1_A2")
helmert

lm(values~phase, data=df, contrasts = list(phase = helmert)) |> coef()

c(Intercept = grand_mean,
  phaseB1vsA1 = (25.9 + -13.7) / 2, 
  phaseA2vsA1_B1 = (14.7  + (-13.7-25.9)/2) / 3,
  phaseB2vsA1_B1_A2 = (27 + (-13.7-25.9-14.7)/3) /4)
```

## Revert Helmert

*Comparing the second to last phase against the mean of all preceding phases*

The Intercept is the grand mean.\
The predictors are the differences from a phase to the mean of the preeding phases.

*Compare phase2 vs. phase 1, phase 3 vs. mean of phase 1 and 2, phase4 vs. mean of phases 1,2,3 (Intercept is overall mean)*

```{r}
revers_helmert <- matrix(c(-1/2, 1/2, 0, 0, -1/3, -1/3, 2/3, 0, -1/4, -1/4, -1/4, 3/4), ncol = 3)
colnames(revers_helmert) <- c("B1vsA1", "A2vsA1_B1", "B2vsA1_B1_A2")
revers_helmert

lm(values~phase, data=df, contrasts = list(phase = revers_helmert)) |> coef()

c(Intercept = grand_mean,
  phaseB1vsA1 = 25.7 - 13.7,
  phaseA2vsA1_B1 = 14.7 - (13.7+25.9)/2,
  phaseB2vsA1_B1_A2 = 27 - (13.7+25.9+14.7)/3)
```

## Cummulative

*Compare mean of second to last phase to mean of preceding phase*

The Intercept is the mean of the first phase.\
The predictors are the differences from a phase to the preeding phase.

```{r}
cumulative <- matrix(c(0,1,1,1, 0,0,1,1, 0,0,0,1), ncol = 3)
colnames(cumulative) <- c("B1vsA1","A2vsB1","B2vsA2")
cumulative

lm(values~phase, data=df, contrasts = list(phase = cumulative)) |> coef()

c(Intercept = 13.7,
  phaseB1vsA1 = 25.9 - 13.7, 
  phaseA2vsB1 = 14.7 - 25.9,
  phaseB2vsA2 = 27 - 14.7)
```

# Contrast computation

## treatment

```{r}
#| collapse: true

(treatment <- contr.treatment(4))
(treatment <-  cbind(1, treatment))

(treatment <- solve(t(treatment)))
(treatment <- treatment[,-1])

colnames(treatment) <- c("B1vsA1","A2vsA1","B2vsA1")

lm(values~phase, data=df, contrasts = list(phase = treatment)) |> summary()
```

## sum

```{r}
(sum <- contr.sum(4))
(sum <-  cbind(1, sum))

(sum <- solve(t(sum)))
(sum <- sum[,-1])
colnames(sum) <- c("A1vsMean","B1vsMean","A2vsMean")


lm(values~phase, data=df, contrasts = list(phase = sum)) |> coef()

c(Intercept = grand_mean,
  phaseA1vsMean = 13.7 - grand_mean, 
  phaseB1vsMean = 25.9 - grand_mean, 
  phaseA2vsMean = 14.7 - grand_mean)
```

