---
title: "Treatment vs. effect contrasts"
author: "Jürgen Wilbert"
date: "`r Sys.Date()`"
format:
  html:
    page-layout: full
    df-print: kable
    toc: true
    toc-location: right
    self-contained: true
categories:
  - regression
  - statistics
  - contrasts
---

### Hi Wolfgang and Michel, here is a simple example to show the differences between treatment and effect contrasts.

# Example dataset

Create a random dataset with criteria `y`, predictors `x1`, `x2` and `gender`.

All variables are correlated.

```{r}
set.seed(1234)
n <- 10000
gender <- rep(0:1, each = n/2)
y <- sample(0:10, n, replace = TRUE) + gender * sample(0:10, n, replace = TRUE)
x1 <- sample(0:10, n, replace = TRUE) + gender * y
x2 <- x1 + sample(0:10, n, replace = TRUE) + gender * y

y <- y + (x1 > median(x1) & x2 > median(x2) & gender == 1) * sample(0:10, n, replace = TRUE)

dat <- data.frame(y = y, x1 = x1, x2 = x2, gender = gender)
```

# Descriptives

```{r}
psych::corr.test(dat)
psych::describe(dat)
```

# Contrasts

The left part of the table is with gender as treatment contrast (0 vs. 1) and the right part with gender as effect contrast (-1 vs. 1)

```{r}
# Gender has values 0 vs. 1 (treatment contrast)
fit1 <- lm(y ~ gender * x1 * x2, data = dat)

# Gender hast -1 vs. 1 (effect contrast)
dat$gender <- car::recode(dat$gender, "0 = -1; 1 = 1")

fit2 <- lm(y ~ gender * x1 * x2, data = dat)

sjPlot::tab_model(fit1, fit2, show.se = TRUE, show.ci = FALSE, col.order = c("est", "se", "std.est", "p"), digits = 4)
```

