---
title: "Analyze experimental data with a multilevel model"
author: "Jürgen Wilbert"
abstract: Material for a brief explanation of how to analyse data from an experimental study with a multilevel model in R.
date: today
format:
  html:
    page-layout: full
    df-print: kable
    toc: true
    toc-location: right
    self-contained: false
    code-fold: true
categories:
  - multilevel
  - R
  - experiments
  - statistics
editor: 
  markdown: 
    wrap: 72
execute: 
  warning: false
  error: false
---

```{r}
#| label: Load data
#| code-summary: Load data

# Load the required library
library(tidyverse)
library(lmerTest)
library(sjPlot)
library(emmeans)
library(ez)

if (!("wmisc" %in% installed.packages())) devtools::install_github("jazznbass/wmisc")
if (!packageDate("wmisc") >= as.Date("2023-12-27")) devtools::install_github("jazznbass/wmisc")
library(wmisc)

dat_items <- readRDS(file.path("data-items.rds"))
dat_items <- dat_items %>%
  mutate(
    group = factor(
      group, levels = c(1, 0), 
      labels = c("Control", "Training")
    ),
    id_subject = sessionToken,
    time = factor(
      run, levels = c("pre", "post"),
      labels = c("Pre", "Post")
    ),
    item_effect = paste0(trend, slope),
    effect = factor(
      item_effect, 
      levels = c("00", "10", "01", "11"),
      labels = c("None", "Trend", "Slope", "Trend+Slope"))
  ) %>%
  arrange(id_subject,curr_timestamp)

dat_subjects <- dat_items %>% 
  filter(question == "effect") %>%
  group_by(id_subject, time, trend, slope, group) %>%
  summarise(
    prop = mean(response, na.rm = TRUE)
  ) %>% 
  mutate(id_subject = factor(id_subject), trend = factor(trend), slope = factor(slope)) %>%
  ungroup()


logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  prob
}

```

# Deisgn

```{r}

dat_items %>% filter(question == "effect") %>%
  group_by(sessionToken, group) %>%
  summarise(n = n()) %>%
  group_by(group) %>%
  summarise(n = n())


```

DV 1: Yes/No ratings of single-case graphs "did the intervention have an effect?"

*DV 2: 1-5 scale "How certain are you about your rating?"*

Between IV 1: Intervention group with training in graphreading vs. controll group without training in graph reading.

Within IV 1: Pre-intervention vs. post-intervention

Within IV 2: Graph shows a trend-effect vs. no trend-effect

Within IV 3: Graph shows a slope-effect vs. no slope-effect

10 graphs per within condition: 2 x 2 x 2 x 10 = 80 Graphs 


# Descriptive analyses


```{r}
dat_items %>%
  filter(question == "effect") %>%
  group_by(group) %>%
  summarise(n = n()/ 80)
```


```{r}
dat_items %>%
  filter(question == "effect") %>%
  group_by(group, effect, time) %>%
  summarise(
    mean_true = round(mean(response, na.rm = TRUE), 2)
  ) %>%
  ungroup() %>%
  pivot_wider(names_from = "time", values_from = "mean_true") %>%
    mutate(
    "Difference" = Post - Pre
    #sdt_category = rep(c("false alarm", "hit"), 4)
  ) %>%
  rename(Condition = group, Effect = effect) %>%
  relocate(Condition, Effect) %>%
  nice_table(
    file = "tab-desc-prop-response.docx", 
    title = "Proportion of graphs rated as showing an intervention effect"
  )

```

# "Traditional" Manova

```{r}
dat_subjects %>% slice(1:20) %>% nice_table()
```


Between subject factor is group and within subject factors are time, trend, and slope.

```{r}
fit <- ezANOVA(
  dat_subjects, 
  wid = id_subject, 
  dv = prop, 
  within = list(time, trend, slope), 
  between = group, 
  type = 3
)

nice_table(fit$ANOVA, decimals = 2)

```

# Analyses with multilevel model

```{r}
dat <- dat_items %>% 
  filter(question == "effect") %>%
  mutate(response = factor(response, labels = c("No", "Yes"))) %>%
  rename(Condition = group, Effect = effect, Time = time)

dat %>% slice(1:30) %>% select(id_subject, Condition, Time, Effect, response) %>% nice_table()
```


1. Variables slope and trend are aggrgated into a new variable "effect" with for levels: "None", "Trend", "Slope", "Trend+Slope"

```{r}
#| tbl-cap: logistic models for assumed true responses (dummy contrasts)



model <- glmer(
  response ~ Time * Condition * Effect + 
             (1|id_subject) + (1|id_subject:Time) + 
             (1|id_subject:Effect), 
  family = binomial,
  nAGQ = 0,
  data = dat, 
  na.action = na.omit
)

sjPlot::tab_model(
  model, 
  #file = "tab-reg-response.doc",
  show.se = FALSE, 
  show.ci = FALSE,
  show.stat = TRUE,
  show.df = FALSE,
  string.se = "se",
  string.est = "OR",
  string.stat = "t"
)

```

## Estimated marginal means from the model

```{r}
marginal_means <- emmeans(model, c("Time", "Effect", "Condition"),
  pbkrtest.limit = 9000,
  lmerTest.limit = 9000
)

means <- summary(marginal_means) |> as.data.frame()
means$probability <- logit2prob(means$emmean)
means$prob.ll <- logit2prob(means$asymp.LCL)
means$prob.ul <- logit2prob(means$asymp.UCL)

table_pre_post <- means %>%
  select(Condition, Effect, Time, probability) %>%
  pivot_wider(names_from = "Time", values_from = "probability") %>%
  mutate(Difference = round(Post-Pre, 2), Pre = round(Pre, 2), Post = round(Post, 2)) 

table_contrast <- marginal_means |> 
  pairs() |> 
  as.data.frame() |> 
    filter(contrast %in% c(
    "Pre None Control - Post None Control",
    "Pre None Training - Post None Training",
    "Pre Trend Control - Post Trend Control",
    "Pre Trend Training - Post Trend Training",
    "Pre Slope Control - Post Slope Control",
    "Pre Slope Training - Post Slope Training",
    "(Pre Trend+Slope Control) - (Post Trend+Slope Control)",
    "(Pre Trend+Slope Training) - (Post Trend+Slope Training)"
  )) %>% 
  mutate(
    across(where(is.numeric), ~round(.x, 2)),
    contrast = case_match(
      contrast,
      "Pre None Control - Post None Control" ~ "Control / none",
      "Pre None Training - Post None Training" ~ "Training / none",
      "Pre Trend Control - Post Trend Control" ~ "Control / trend",
      "Pre Trend Training - Post Trend Training" ~ "Training / trend",
      "Pre Slope Control - Post Slope Control" ~ "Control / slope",
      "Pre Slope Training - Post Slope Training" ~ "Training / slope",
      "(Pre Trend+Slope Control) - (Post Trend+Slope Control)" ~ "Control / trend+slope",
      "(Pre Trend+Slope Training) - (Post Trend+Slope Training)" ~ "Training trend+slope"
    )
  ) %>%
  rename("z ratio" = z.ratio, p = p.value) %>% 
  select(-1, -estimate, -df)

table_pre_post <- cbind(table_pre_post, table_contrast)
  nice_table(
    table_pre_post,
    file = "tab-marginal-means-response.docx",
    title = "Pre/post post-hoc contrasts of proportions of graphs rated as showing an intervention effect"
  )

```

```{r}
means %>% 
  mutate(Percentage = probability * 100) %>%
  ggplot(aes(x = Time, y = Percentage)) +
  geom_line(
    aes(color = Effect, group = Effect, linetype = Effect),
    position = position_dodge(0.2)
  ) + 
  geom_point(
    aes(group = Effect),
    position = position_dodge(0.2)
  ) + 
  geom_errorbar(
    aes(ymin = prob.ll * 100, ymax = prob.ul * 100), 
    width = 0.2,
    position = position_dodge2(0.2)
  ) +  
  facet_grid(cols = vars(Condition)) +
  ylim(c(0,100)) +
  theme(
    panel.background = element_rect(fill = "white"),
    legend.key  = element_rect(fill = "white"),
    axis.line.x = element_line(colour = "black", linewidth = 1),
    axis.line.y = element_line(colour = "black", linewidth = 1)
  ) + 
  xlab("Time") +
  ggtitle("Estimated marginal means of response that an effect exists")


```

